---
output: github_document
---

# Cartography in R: and introduction

The vector data model represents the world using points, lines and polygons.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Starting with `{sf}`

`{sf}` stands for simple features, which refers to a formal standard (ISO 19125-1:2004) that describes how objects in the real world can be represented in computers, with emphasis on the spatial geometry of these objects. It also describes how such objects can be stored in and retrieved from databases, and which geometrical operations should be defined for them.
As we will see later, the `{sf}` library works very well with the `{tidyverse}`

```{r message = FALSE}
# install.packages(c("tidyverse", "sf", "ggrepel",
#                    "tmap", "tmaptools"))


library(tidyverse)
```


```{r }
library(sf) # the library for spatial data visualisation
```

`{sf}` functions are all consistent: they all start with `st_*`, which stands for spatio temporal.

Let's start by creating a point in space.

```{r}
# creating our first spatial object, a point in space
unimail_point <- st_point(x = c(46.1952452857602, 6.14051554056255))

unimail_point 

class(unimail_point)
```

```{r}

# creating  a more familiar tibble, with latitude and longitude coordinates
# each location is a vertex of unimail

unimail_df <- tribble(
  ~location, ~lat, ~lon,
  "a", 46.19377633846322, 6.1399410679254744,
  "b", 46.19491952134567, 6.141224481424885,
  "c", 46.194671442623196, 6.138677026705302,
  "d", 46.195630227161544, 6.139834520389675
  )

unimail_df 
```

```{r}
# we transform the unimail_df into a sf data.frame

unimail <- unimail_df %>% 
  st_as_sf(coords=c("lon", "lat"), #selecting the variables with the coordinates
           crs = "EPSG:4326") # selecting the projection


unimail
```


```{r}
plot(unimail)
```


```{r}
# from a series of point, we actually want to create a polygon

unimail_polygon <- unimail %>%
  mutate(location_name = "unimail") %>% # creating a grouping variable
  group_by(location_name) %>%
  dplyr::summarise() %>%
  st_cast("POLYGON") %>% # will cast the points to a polygon
  st_convex_hull() # makes sure the polygon is not convex

unimail_polygon
```

```{r}
plot(unimail_polygon)
```



```{r}
class(unimail_polygon)

unimail_polygon
```

## From sfg (simple feature geometry) to sfc (simple feature column) to sf data frame

Why use this data type and `{sf}`

- Fast reading and writing of data
- **visualisation**: enhanced plotting performance and new `geom_sf()` for `{ggplot}.` 
-  **transferability**`sf data.frames` can be treated as data frames in most operations
-  **uniformity**: `{sf}` functions can be combined using %>% operator and works
-  works well with the tidyverse collection of R packages
- **consistency**sf function names are relatively consistent and intuitive (all begin with st_)

## Properties of `sf data.frame`s

- geometries are sticky

```{r}

unimail_polygon %>% 
  select(-geometry)

unimail_polygon %>% 
  select(location_name)

unimail_polygon[,"location_name"]

```

Only way to get rid of those is using `st_drop_geometry()`.

```{r}
st_drop_geometry(unimail_polygon)
```


## Waste bins in Geneva

Data downloaded from the Système d'information du territoire à Genève (SITG).

`st_read()` is used to read in geometries, that are stored in shapefiles.


```{r}
# locations of geneva bins
# loading a shapefile

bins <- st_read("data/SHP_VDG_CORBEILLES_DECHETS/VDG_CORBEILLES_DECHETS.shp") 
```

The data is although projectes using swiss mercator: "CH1903+ / LV95", we want to use although the same projection as the `unimail_polygon`, which is in the 
projection used in google maps.


```{r}
bins <- bins %>% 
  st_transform(crs = "EPSG:4326") #apply transformation

```


```{r}
# shapes for the roads
roads <- st_read("data/SHP_GMO_GRAPHE_ROUTIER/GMO_GRAPHE_ROUTIER.shp") %>% 
  st_transform(crs = "EPSG:4326")

# shapes for the water
water <- st_read("data/SHP_GEO_LAC/GEO_LAC.shp") %>% 
  st_transform(crs = "EPSG:4326")
```

## Plotting geometries with `{ggplot2}`

`geom_sf()` is our new ggplot2 friend

```{r}
ggplot()+
  geom_sf(
    data =water,
    fill = "lightblue"
    )
```


```{r}
ggplot()+
  geom_sf(
    data =water,
    fill = "lightblue"
    )+
  geom_sf(
    data = roads,
    color = "#fde293")
```

```{r}
ggplot()+
  geom_sf(
    data =water,
    fill = "lightblue"
    )+
  geom_sf(
    data = roads,
    color = "#fde293") +
  theme_void()+ # adding theme void
  coord_sf(ylim = c(46.193854, 46.205765),
           xlim = c(6.134706, 6.151571)) # zooming in
```


```{r}
# plot the data 
ggplot()+
  geom_sf(
    data =water,
    fill = "lightblue"
    )+
  geom_sf(
    data = roads,
    color = "#fde293") +
  geom_sf(data =unimail_polygon,  # adding polys
          fill = "yellow") +
  geom_sf(data = bins, # and the bins
          alpha = 0.1) +
  coord_sf(ylim = c(46.193854, 46.205765),
           xlim = c(6.134706, 6.151571)) +
  labs(title = "Geneva",
       caption = "Source: Système d'information \n du territoire à Genève (SITG)")+
  theme_void(base_size = 10)

```



## Manipulating geometries

```{r}
st_intersection(unimail_polygon, bins)
```

```{r}
ggplot()+
  geom_sf(
    data =water,
    fill = "lightblue"
    )+
  geom_sf(
    data = roads,
    color = "#fde293") +
  geom_sf(data =unimail_polygon,  # adding polys
          fill = "yellow") +
  geom_sf(data = st_intersection(unimail_polygon, bins)) + #only unimail bin
  coord_sf(ylim = c(46.193854, 46.205765),
           xlim = c(6.134706, 6.151571)) +
  labs(title = "Geneva",
       caption = "Source: Système d'information \n du territoire à Genève (SITG)")+
  theme_void(base_size = 10)
```

Let's look a bit further away from the polygon


```{r}
buffered_polys <-  st_buffer(unimail_polygon, 
                             dist = 0.001) # distance is in degrees..

ggplot()+
  geom_sf(
    data = buffered_polys,
    fill = "lightblue"
    )+
  geom_sf(
    data = unimail_polygon,
    color = "#fde293")
```


```{r}
st_intersection(buffered_polys, bins)
```


```{r}

ggplot()+
  geom_sf(
    data =water,
    fill = "lightblue"
    )+
  geom_sf(
    data = roads,
    color = "#fde293") +
  geom_sf(data =unimail_polygon, 
          fill = "yellow") +
  geom_sf(data = st_intersection(buffered_polys, bins))+ # bins close to unimail 
  coord_sf(ylim = c(46.193854, 46.205765),
           xlim = c(6.134706, 6.151571)) +
  labs(title = "Geneva",
       subtitle= glue::glue("Near UNIMAIL there are {nrow(st_intersection(buffered_polys, bins))} bins "),
       caption = "Source: Système d'information \n du territoire à Genève (SITG)")+
  theme_void(base_size = 10)

```

## Overview presentation

## Making thematic maps (choropleth maps) of COVID-19 data


```{r}
#importing COVID data
swiss_covid_latest <- read_csv("data/latest_swiss_data.csv")

glimpse(swiss_covid_latest)
```

### Importing the shapefiles

```{r}
canton_shape <- st_read("data/SHAPEFILE_LV03_LN02/swissBOUNDARIES3D_1_3_TLM_KANTONSGEBIET.shp")

glimpse(canton_shape)
```


```{r}
codes <- read_csv("data/canton_codes.csv")

codes
```


```{r}
canton_shape <- canton_shape %>% 
  left_join(codes, c("KANTONSNUM" = "code_num")) %>% 
  left_join(swiss_covid_latest, c("code"= "abbreviation_canton_and_fl"))

glimpse(canton_shape)
```


```{r}

#removing non useful columns

canton_shape <- canton_shape %>% 
  select(code, incidence)

glimpse(canton_shape)
```

### The `ggplot2` way

```{r}
ggplot(data= canton_shape, 
       aes(fill = incidence))+
  geom_sf() +
  theme_void()
```

I will now divide the continuous variable into categories.
I will use quantiles, as usually this way allows for colors, and therefore categories, to be evenly distributed in our map.


```{r}
quantile_vec <- quantile(canton_shape$incidence, 
                         na.rm = TRUE, 
                         probs = seq(0, 1, 0.2))

quantile_vec
```


```{r}
# I create a label to be used then in the plot

labels <- tibble(
  lab1 = quantile_vec,
  lab2 = c(quantile_vec[2:length(quantile_vec)], NA)) %>%
  slice(1:n() - 1) %>% 
  mutate_all(round, digits = 0) %>% 
  mutate(labs = paste(lab1, lab2, sep = " -"))

labels
```


```{r}

canton_shape <- canton_shape %>%
  mutate(incidence_cut = cut(incidence, # cutting the incidence
    breaks = quantile_vec,
    labels = labels %>% pull(labs),
    include.lowest = TRUE, # includes the lowest 'breaks' value
    ordered_result = TRUE, # orders the factors/bins
  ))

glimpse(canton_shape)
```

### Plotting the data, the ggplot way


```{r}

ggplot(data = canton_shape) +
  geom_sf(aes(fill = incidence_cut)) +
  rcartocolor::scale_fill_carto_d(type = "quantitative", 
                                  palette = "BurgYl")+
  theme_void() +
  labs(fill = "",
       title = paste0("Cumulative confirmed ", 
                      "cases in Switzerland (per 100'000 population)"),
       caption = str_glue("source: OFSP | updated {unique(canton_shape$date)}"))
```

```{r}

## Add labels to show the incidence by canton

ggplot(data = canton_shape) +
  geom_sf(aes(fill = incidence_cut)) +
  rcartocolor::scale_fill_carto_d(type = "quantitative", 
                                  palette = "BurgYl")+
  ggrepel::geom_label_repel(
    data = canton_shape,
    aes(label = paste0(code,":",round(incidence, digits = 0)), 
        geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0.2,
    colour = "#541f3f",
    size = 3,
    segment.alpha = 0.5
  ) +
  theme_void() +
  labs(fill = "",
       title = paste0("Cumulative confirmed ", 
                      "cases in Switzerland (per 100'000 population)"),
       caption = str_glue("source: OFSP | updated {unique(canton_shape$date)}"))

```




### The `tmap` way

```{r}
# install.packages(tmap)
# install.packages(tmaptools)
library(tmap) # library for thematic maps


tm_shape(canton_shape) + 
  tm_polygons(col = "incidence", 
              style = "quantile", 
              palette = "-viridis")


```


```{r}
osmbb <-  tmaptools::bb(canton_shape,
                        ext=1.1 )# to enlarge a bit the bounding box

osmtiles <- tmaptools::read_osm(osmbb, type="esri-topo")

tm_shape(osmtiles) + 
  tm_rgb() +
  tm_shape(canton_shape) + 
  tm_polygons(col = "incidence", 
              style = "quantile", 
              palette = "-viridis") +
  tm_layout(frame = FALSE)
```

To make your map interactive, use `tmap_mode("view")` before plotting it

```{r}
# tmap_mode("view") # make interactive
```



## Dive deeper

- [Geocomputation with R](https://geocompr.robinlovelace.net)
- [Spatial Data Science](https://keen-swartz-3146c4.netlify.app)
- [An R spatial tutorial](https://github.com/statnmap/user2020_rspatial_tutorial/blob/master/slides_user2020_rspatial.pdf)
- [examples from 30 day map challenge](https://github.com/Nowosad/rspatial-map-challenge-2020) 
- on Twitter check #rspatial



