---
output: github_document
---

# Cartography in R: an introduction to vector data

This tutorial will only touch vector data model, that represents the world using points, lines and polygons. The go-to library in `R` to work with vector data is `{sf}`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Starting with `{sf}`: creating a `{sf}` objects

`{sf}` stands for simple features, which refers to a formal standard (ISO 19125-1:2004) that describes how objects are represented in computer. It also describes how such objects can be stored or retrieved from databases, and which geometrical operations can be defined for them.

As we will see, the `{sf}` library works very well with the `{tidyverse}`.

```{r message = FALSE}
# The packages to install for today
# install.packages(c("tidyverse", "sf", "ggrepel",
#                    "tmap", "tmaptools"))


library(tidyverse)
```

Useful to know: the `{sf}` library relies on external dependencies (GEOS, GDAL and PROJ). It is good to be aware of this as updates in those libraries can result in broken R code. Beware of this.

Let's load the `{sf}` library.

```{r }
library(sf) # the library for spatial data visualisation
```

`{sf}` functions are all consistent: they all start with `st_*`, which stands for `s`patio `t`emporal.

Let's start by creating a point in space, which corresponds to the centre of Geneva University, UniMail.

```{r}
# creating our first spatial object, a point in space
unimail_point <- st_point(x = c(46.1952452857602, 6.14051554056255))

# check it out how it looks like and its class
unimail_point 

class(unimail_point)
```

Since UniMail is actually best represented by a polygon than by one point, let's see how we can build a polygon for it.

```{r}

# creating  a more familiar tibble, with latitude and longitude coordinates
# each location is a vertex of unimail building

unimail_df <- tribble(
  ~location, ~lat, ~lon,
  "a", 46.19377633846322, 6.1399410679254744,
  "b", 46.19491952134567, 6.141224481424885,
  "c", 46.194671442623196, 6.138677026705302,
  "d", 46.195630227161544, 6.139834520389675
)

unimail_df
```

```{r}
# we transform the unimail_df into a sf data.frame

unimail <- unimail_df %>%
  st_as_sf(
    coords = c("lon", "lat"), # selecting the variables with the coordinates
    crs = "EPSG:4326" # selecting the projection
  ) 

# check out how the coordinates are stored
# a new nested column has been created which is called geometry

unimail
```

```{r}
class(unimail)
```

From a simple tibble, we have created an `sf data.frame`, which has a `geometry` column that is a list column containing the geometries we want. The geometry column is an `sfc`, a `simple feature column`.

```{r}
plot(unimail)
```

We now have 4 points, let's create a polygon.

```{r}

# from a series of point, we actually want to create a polygon

unimail_polygon <- unimail %>%
  mutate(location_name = "unimail") %>% # creating a grouping variable
  group_by(location_name) %>%
  dplyr::summarise() %>%
  st_cast("POLYGON") %>% # will cast the points to a polygon
  st_convex_hull() # makes sure the polygon is not convex, try to remove it to see


unimail_polygon
```

```{r}
plot(unimail_polygon)
```

```{r}
class(unimail_polygon)

unimail_polygon
```

Why use this data type and `{sf}`

-   Fast reading and writing of data
-   **visualisation**: enhanced plotting performance (`plot()`) and new `geom_sf()` for `{ggplot}.`
-   **transferability**: `sf data.frames` can be treated as data frames in most operations
-   **compatibility**: `{sf}` functions can be combined using `%>%` operator
-   **compatibility**: works well with the tidyverse collection of R packages
-   **consistency**: `{sf}` function names are relatively consistent and intuitive (all begin with `st_`)

## Properties of `sf data.frame`s

-   geometries are sticky!

```{r}
# Try these commands, they will not remove the geometry column

unimail_polygon %>%
  select(-geometry)

unimail_polygon %>%
  select(location_name)

unimail_polygon[, "location_name"]
```

Only way to get rid of those is using `st_drop_geometry()`.

```{r}
st_drop_geometry(unimail_polygon)
```

## Case Study 1: Checking waste bins close to UniMail building

Data downloaded from the Système d'information du territoire à Genève (SITG).

`st_read()` is used to read in geometries, that are stored in shapefiles.

```{r}
# locations of geneva bins
# loading a shapefile

bins <- st_read("data/SHP_VDG_CORBEILLES_DECHETS/VDG_CORBEILLES_DECHETS.shp")
```

The data is although projected using Swiss Mercator: "CH1903+ / LV95", we want to use although the same projection as the `unimail_polygon`, which is in the same projection used in Google Maps.

```{r}
bins <- bins %>%
  st_transform(crs = "EPSG:4326") # apply projection transformation 
```

```{r}
# shapes for the roads: a collection of linestrings
roads <- st_read("data/SHP_GMO_GRAPHE_ROUTIER/GMO_GRAPHE_ROUTIER.shp") %>%
  st_transform(crs = "EPSG:4326")

# shapes for the water: a collection of polygons
water <- st_read("data/SHP_GEO_LAC/GEO_LAC.shp") %>%
  st_transform(crs = "EPSG:4326")
```

### Plotting geometries with `{ggplot2}`

`geom_sf()` is our new `{ggplot2}` friend. It will automatically check the type of geometry it needs to plot.

```{r}
ggplot() +
  geom_sf(
    data = water,
    fill = "lightblue"
  )
```

For plotting spatial data we reason in the same way we do with `{ggplot2}`, we add layer by layer and the layer order matters!

```{r}
ggplot() +
  geom_sf(
    data = water,
    fill = "lightblue"
  ) +
  geom_sf(
    data = roads,
    color = "#fde293"
  )
```

```{r}
ggplot() +
  geom_sf(
    data = water,
    fill = "lightblue"
  ) +
  geom_sf(
    data = roads,
    color = "#fde293"
  ) +
  theme_void() + # adding theme void
  coord_sf(
    ylim = c(46.193854, 46.205765),
    xlim = c(6.134706, 6.151571)
  ) # zooming in
```

```{r}
# plot the data
ggplot() +
  geom_sf(
    data = water,
    fill = "lightblue"
  ) +
  geom_sf(
    data = roads,
    color = "#fde293"
  ) +
  geom_sf(
    data = unimail_polygon, # adding polys
    fill = "yellow"
  ) +
  geom_sf(
    data = bins, # and the bins
    alpha = 0.1
  ) +
  coord_sf(
    ylim = c(46.193854, 46.205765),
    xlim = c(6.134706, 6.151571)
  ) +
  labs(
    title = "Geneva",
    caption = "Source: Système d'information \n du territoire à Genève (SITG)"
  ) +
  theme_void(base_size = 10)
```

### Manipulating the unimail polygon

The advantage of using `{sf}` objects to deal with spatial data, is that we can do spatial operations with them, in a pretty simple way. Let's find the city trash bins that intersect with our UniMail polygon. How many are there?

```{r}
st_intersection(unimail_polygon, bins)
```

Just 1 trash bin, let's plot it:

```{r}
ggplot() +
  geom_sf(
    data = water,
    fill = "lightblue"
  ) +
  geom_sf(
    data = roads,
    color = "#fde293"
  ) +
  geom_sf(
    data = unimail_polygon, # adding polys
    fill = "yellow"
  ) +
  geom_sf(data = st_intersection(unimail_polygon, bins)) + # only unimail bin
  coord_sf(
    ylim = c(46.193854, 46.205765),
    xlim = c(6.134706, 6.151571)
  ) +
  labs(
    title = "Geneva",
    caption = "Source: Système d'information \n du territoire à Genève (SITG)"
  ) +
  theme_void(base_size = 10)
```

Let's look a bit further away from the polygon. To do this, we need to create a buffer around the UniMail polygon and use this buffered zone as intersect.

```{r}
buffered_polys <- st_buffer(unimail_polygon,
  dist = 0.001
) # distance is in degrees..

ggplot() +
  geom_sf(
    data = buffered_polys,
    fill = "lightblue"
  ) +
  geom_sf(
    data = unimail_polygon,
    color = "#fde293"
  )
```

Then we can repeat the same operation, but you will see now that there are many more trash bins that intersect with the buffered geometry.

```{r}
st_intersection(buffered_polys, bins)
```

```{r}

ggplot() +
  geom_sf(
    data = water,
    fill = "lightblue"
  ) +
  geom_sf(
    data = roads,
    color = "#fde293"
  ) +
  geom_sf(
    data = unimail_polygon,
    fill = "yellow"
  ) +
  geom_sf(data = st_intersection(buffered_polys, bins)) + # bins close to unimail
  coord_sf(
    ylim = c(46.193854, 46.205765),
    xlim = c(6.134706, 6.151571)
  ) +
  labs(
    title = "Geneva",
    subtitle = glue::glue("Near UNIMAIL there are {nrow(st_intersection(buffered_polys, bins))} bins "),
    caption = "Source: Système d'information \n du territoire à Genève (SITG)"
  ) +
  theme_void(base_size = 10)
```

## Case Study 2: Making thematic maps (choropleth maps) of COVID-19 data

### Importing the data

```{r}
# importing COVID data
swiss_covid_latest <- read_csv("data/latest_swiss_data.csv")

glimpse(swiss_covid_latest)
```

```{r}
canton_shape <- st_read("data/SHAPEFILE_LV03_LN02/swissBOUNDARIES3D_1_3_TLM_KANTONSGEBIET.shp") # shapefiles for swiss cantons

glimpse(canton_shape)
```

```{r}
# will use this dataset to be able to merge swiss_covid_latest with
# canton_shapes.

codes <- read_csv("data/canton_codes.csv")

codes
```

```{r}
canton_shape <- canton_shape %>%
  left_join(codes, c("KANTONSNUM" = "code_num")) %>%
  left_join(swiss_covid_latest, c("code" = "abbreviation_canton_and_fl"))

glimpse(canton_shape)
```

```{r}
# removing non useful columns

canton_shape <- canton_shape %>%
  select(code, incidence)

glimpse(canton_shape)
```

### Plotting it the `ggplot2` way

```{r}
ggplot(
  data = canton_shape,
  aes(fill = incidence)
) +
  geom_sf() +
  theme_void()
```

You may have noticed from the amount of thematic maps that you have probably digested in 2020 (COVID-19 and US elections to name a few big topics) that continuous colour scales are divided in categories, so we need to work with a categorical variable. I will now divide the continuous variable into categories. I will do this "manually" first computing the quantiles, as usually this way allows for colours, and therefore categories, to be evenly distributed in our map.

> They would be actually evenly distributed if the areas were all of the same size. This is one of the main flaws of thematic maps, as they tend to magnify big areas. You have probably heard the motto "Land doesn't vote, people do" that was used to highlight the flaws of this type of map to show votes for the US elections. This flaw was also used by Trump to justify the idea that 2016 election was a landslide...

```{r}
quantile_vec <- quantile(canton_shape$incidence,
  na.rm = TRUE,
  probs = seq(0, 1, 0.2)
)

quantile_vec
```

```{r}
# I create a label to be used then in the plot

labels <- tibble(
  lab1 = quantile_vec,
  lab2 = c(quantile_vec[2:length(quantile_vec)], NA)
) %>%
  slice(1:n() - 1) %>%
  mutate_all(round, digits = 0) %>%
  mutate(labs = paste(lab1, lab2, sep = " -"))

labels
```

```{r}

canton_shape <- canton_shape %>%
  mutate(incidence_cut = cut(incidence, # cutting the incidence
    breaks = quantile_vec,
    labels = labels %>% pull(labs),
    include.lowest = TRUE, # includes the lowest 'breaks' value
    ordered_result = TRUE, # orders the factors/bins
  ))

glimpse(canton_shape)
```

`incidence_cut` is the ordered categorical variable I have created and that I will use to map the `fill` argument to.

### Plotting the data, the `{ggplot}` way

```{r}

ggplot(data = canton_shape) +
  geom_sf(aes(fill = incidence_cut)) +
  rcartocolor::scale_fill_carto_d(
    type = "quantitative",
    palette = "BurgYl"
  ) +
  theme_void() +
  labs(
    fill = "",
    title = paste0(
      "Cumulative confirmed ",
      "cases in Switzerland (per 100'000 population)"
    ),
    caption = str_glue("source: OFSP | updated {unique(canton_shape$date)}")
  )
```

```{r}

## Add labels to show the incidence by canton

ggplot(data = canton_shape) +
  geom_sf(aes(fill = incidence_cut)) +
  rcartocolor::scale_fill_carto_d(
    type = "quantitative",
    palette = "BurgYl"
  ) +
  ggrepel::geom_label_repel(
    data = canton_shape,
    aes(
      label = paste0(code, ":", round(incidence, digits = 0)),
      geometry = geometry
    ),
    stat = "sf_coordinates",
    min.segment.length = 0.2,
    colour = "#541f3f",
    size = 3,
    segment.alpha = 0.5
  ) +
  theme_void() +
  labs(
    fill = "",
    title = paste0(
      "Cumulative confirmed ",
      "cases in Switzerland (per 100'000 population)"
    ),
    caption = str_glue("source: OFSP | updated {unique(canton_shape$date)}")
  )
```

### Exercise

You might have noticed that there are more labels than what is needed. This results from the fact that we are using high resolution shapes, which result in the fact that each canton might be described by multiple polygons (e.g. Geneva has a piece of land that is in the middle of Canton Vaud). Try to make a new variable, that will contain only one label per canton, for plotting purposes.

### The `tmap` way

We have seen the `{ggplot2}` way of creating thematic maps. I tend to use and teach mostly `{ggplot2}` because of the transferability of concepts and because of the possibilities to use nice extension packages like `{cowplot}`, `{patchwork}`, `{gganimate}` and `{ggtext}`. There is also another reason: once you have learnt how to work with `{ggplot2}`, `{tmap}` (and `{leaflet}` for interactive maps) will be rather intuitive.

`{tmap}` is indeed a package meant for plotting thematic maps and its conciseness is quite unbeatable. Look at the following code. You can see that with the `style` argument you can directly choose how to set categories for the continuous numerical variable. Quite convenient indeed...

```{r}
# install.packages(tmap)
# install.packages(tmaptools)

library(tmap) # library for thematic maps

tm_shape(canton_shape) +
  tm_polygons(
    col = "incidence",
    style = "quantile", # it divides the continuous colour scale using quantiles
    palette = "YlOrBr" # using an rcolorbrewer here
  )
```

`{tmap}` has actually a lot of other nice cards to play. First, we can get the background tiles to plot our map on top. In this case, I will download the base tile from open street map, of the type `esri-topo` and then plot the choropleth (i.e. thematic map) on top of it.

```{r}
# first I compute a bounding box around the cantons area
osmbb <- tmaptools::bb(canton_shape,
  ext = 1.1 # to enlarge a bit the bounding box
) 

# then I use the bounding box to download only that tile
osmtiles <- tmaptools::read_osm(osmbb, type = "esri-topo")

# now I can plot the choropleth on top of osmtiles
tm_shape(osmtiles) +
  tm_rgb() + 
  tm_shape(canton_shape) +
  tm_polygons(
    col = "incidence",
    style = "quantile",
    palette = "YlOrBr"
  ) +
  tm_layout(frame = FALSE)
```

Second, we can easily make your map interactive, use `tmap_mode("view")` before plotting it. You can try running this command and then rerun the previous code. The shapes will be then plotted on top of a standard ESRI map.

```{r}
#tmap_mode("view") # make interactive
```

> If you are getting errors because of the `{rJava}` installation on your macOS, you can check the following resources:

-   [Installing RJava on MacOs](https://zhiyzuo.github.io/installation-rJava/)

If `cask java8` is not possible, use `brew cask install homebrew/cask-versions/adoptopenjdk8` as pointed [here](https://github.com/dennisausbremen/macOStrap/issues/4).

Another good resouce:

-   <https://github.com/Utah-Data-Science/Home_repo/wiki/Getting-R-to-use-the-correct-Java-version>

## Other packages for vector data

-   `{stars}`: This packages is an enhancement of `{sf}` that also works with rasters and it is meant for multidimensional datasets. It draws from the idea of data cubes, and it is helpful when dealing with time series spatial data. The problem with time series is that, if we work with tidy data, we will repeat geometries multiple times, making the data very large. This [package](https://r-spatial.github.io/stars/) is meant to deal exactly with this issue.

-   `{leaflet}`: the go-to package for interactive maps. Take a look at the documentation [here](https://rstudio.github.io/leaflet/) to find more about it.

## More resources

-   [Geocomputation with R](https://geocompr.robinlovelace.net)
-   [Spatial Data Science](https://keen-swartz-3146c4.netlify.app)
-   [An R spatial tutorial](https://github.com/statnmap/user2020_rspatial_tutorial/blob/master/slides_user2020_rspatial.pdf)
-   [examples from 30 day map challenge](https://github.com/Nowosad/rspatial-map-challenge-2020)
-   on Twitter check \#rspatial
